Elephant Seal Age JAE Intraspecific Variation

### 2024-05-23 paper/paperv3.qmd
pup surv/recr pre-/post-senescence

We found that neither pup survival nor recruitment is significantly affected by maternal age (@fig-repro_age).

@@

```{r}
#| label: fig-repro_age
#| fig-cap: "Pup survival and recruitment are not significantly affected by maternal age. Points are mean values for maternal age. Error bars show SE."
#| fig-width: 5.5

@@

surv_recruit_data <- sealdat %>% 
  drop_na(pupID) %>% 
  filter(pupsex == "F",
         pupyearborn < 2016) %>% 
  mutate(pup_survived_dbl = ifelse(pup_survived, 1, 0)) %>% 
  group_by(age) %>% 
  mutate(n_pups = n()) %>% 
  ungroup() %>% 
  filter(n_pups >= 5)
pup_surv_model <- glmer(
  pup_survived_dbl ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year),
  data = surv_recruit_data, 
  family = "binomial"
)
summary(pup_surv_model)
pup_surv_pred <- ggpredict(pup_surv_model, terms = c("age10 [all]")) %>% 
  as_tibble() %>% 
  transmute(age = x * 10, 
            metric = "surv", 
            mean = predicted, 
            mean_lwr = conf.low, 
            mean_upr = conf.high)
pup_recruit_model <- glmer(
  puprecruited ~ poly(age10, 2, raw = TRUE) + (1 | animalID) + (1 | year),
  data = surv_recruit_data, 
  family = "binomial"
)
summary(pup_recruit_model)
pup_recruit_pred <- ggpredict(pup_recruit_model, terms = c("age10 [all]")) %>% 
  as_tibble() %>% 
  transmute(age = x * 10, 
            metric = "recruit", 
            mean = predicted, 
            mean_lwr = conf.low, 
            mean_upr = conf.high)
surv_recruit_pred <- rbind(pup_surv_pred, pup_recruit_pred) %>% 
  mutate(metric = factor(metric, 
                         levels = c("surv", "recruit"), 
                         labels = c("Survival", "Recruitment")))

surv_recruit_summary <- surv_recruit_data %>% 
  group_by(age) %>%
  summarize(n_pups = n(),
            n_surv = sum(pup_survived),
            mean_surv = n_surv / n_pups,  
            se_surv = (mean_surv * (1 - mean_surv) / n_pups)^0.5,
            n_recruit = sum(puprecruited), 
            mean_recruit = n_recruit / n_pups, 
            se_recruit = (mean_recruit * (1 - mean_recruit) / n_pups)^0.5)  %>%  
  pivot_longer(cols = c(mean_surv, se_surv, mean_recruit, se_recruit), 
               names_to = c("stat", "metric"), 
               names_pattern = "(.*)_(.*)") %>% 
  pivot_wider(names_from = "stat", 
              values_from = "value") %>% 
  mutate(low = mean - se, 
         high = mean + se, 
         age = age + ifelse(metric == "surv", -0.1, 0.1), 
         metric = factor(metric, 
                         levels = c("surv", "recruit"), 
                         labels = c("Survival", "Recruitment")))

surv_recruit_summary %>% 
  ggplot(aes(x = age, y = mean, color = metric, fill = metric)) + 
  geom_ribbon(aes(ymin = mean_lwr, ymax = mean_upr, alpha = metric),
              surv_recruit_pred,
              color = NA) +
  geom_pointrange(aes(ymin = low, ymax = high)) +
  geom_line(aes(linetype = metric),
            surv_recruit_pred) +
  scale_y_continuous(labels = scales::percent) + 
  scale_color_manual(values = c("#829BD1", "#E9B668")) + 
  scale_fill_manual(values = c("#829BD1", "#E9B668")) + 
  scale_linetype_manual(values = c("dashed", "dashed")) +
  scale_alpha_manual(values = c(0.25, 0.1)) +
  expand_limits(y = c(0, 0.5)) +
  labs(x = "Maternal Age (Years)", y = "Pup Survival or Recruitment") +
  theme_few() + 
  theme(legend.position = c(0.1,0.97), 
        legend.justification = c(0,1), 
        legend.title = element_blank(), 
        legend.direction = "horizontal", 
        axis.title = element_text(size = 13))

#Why is recruitment higher than survival in some points? Add the quadratic model? Reverse order of legend in the figure
#caption - error bars show SE. Points are mean values for maternal age. Make sure that language in the caption matches both the model and the data 
```

>>>>

We found that neither pup survival nor recruitment is significantly affected by maternal age (@fig-repro-age).

@@

```{r}
#| label: repro-age

# Isolate pup survival/recruitment 
surv_recruit_data <- sealdat %>% 
  drop_na(pupID) %>% 
  # Only pups born 2016 or earlier (otherwise survival biiased)
  filter(pupyearborn <= 2016) %>% 
  mutate(pup_survived_int = ifelse(pup_survived, 1L, 0L)) %>% 
  # Only keep maternal age classes with 5+ births, which drops moms 19+
  group_by(age) %>% 
  mutate(n_pups = n()) %>% 
  ungroup() %>%
  filter(n_pups >= 5)

# Fit pup *survival* models pre- and post-senescence
pup_surv_model1 <- glmer(
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age < age_senesce),
  family = "binomial"
)
pup_surv_model2 <- glmer(
  pup_survived_int ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, age >= age_senesce),
  family = "binomial"
)

# Fit pup *recruitment* models pre- and post-senescence
pup_recr_model1 <- glmer(
  puprecruited ~ age10 + (1 | animalID) + (1 | year),
  # For recruitment, ONLY female pups
  data = filter(surv_recruit_data, 
                age < age_senesce,
                pupsex == "F"),
  family = "binomial"
)
pup_recr_model2 <- glmer(
  puprecruited ~ age10 + (1 | animalID) + (1 | year),
  data = filter(surv_recruit_data, 
                age >= age_senesce,
                pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer="bobyqa")
)

# Summary table
tibble(
  response = rep(c("Survival", "Recruitment"), each = 2),
  n = map_dbl(
    list(pup_surv_model1@frame, 
         pup_surv_model2@frame, 
         pup_recr_model1@frame, 
         pup_recr_model2@frame),
    nrow
  ),
  model = rep(c("Pre-senescence", "Post-senescence"), 2),
  age_coef = c(fixef(pup_surv_model1)["age10"],
               fixef(pup_surv_model2)["age10"],
               fixef(pup_recr_model1)["age10"],
               fixef(pup_recr_model2)["age10"]),
  age_coef_se = c(summary(pup_surv_model1)$coef["age10", "Std. Error"],
                  summary(pup_surv_model2)$coef["age10", "Std. Error"],
                  summary(pup_recr_model1)$coef["age10", "Std. Error"],
                  summary(pup_recr_model2)$coef["age10", "Std. Error"]),
  `p(coef<0)` = c(pnorm(summary(pup_surv_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_surv_model2)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model1)$coef["age10", "z value"]),
                  pnorm(summary(pup_recr_model2)$coef["age10", "z value"]))
) %>% 
  mutate(across(where(is.numeric), \(x) signif(x, 3)))

```{r}
#| label: fig-repro-age
#| fig-cap: "Pup survival and recruitment are not significantly affected by maternal age. Points are mean values for maternal age. Error bars show SE."
#| fig-width: 5.5

@@

# Pup survival 
surv_pred <- rbind(
  # Pre-senescence
  ggpredict(
    pup_surv_model1,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "young"),
  # Post-senescence
  ggpredict(
    pup_surv_model2,
    terms = c("age10 [all]"),
    type = "fixed",
    interval = "confidence"
  ) %>% 
    as_tibble() %>% 
    mutate(age = x * 10,
           age_class = "old")
)

repro_raw <- surv_recruit_data %>% 
  mutate(pupsex = factor(pupsex, labels = c("Female", "Male"))) %>% 
  group_by(age, pupsex) %>% 
  summarize(n_born = n(), 
            n_surv = sum(pup_survived_int),
            perc_surv = n_surv / n_born, 
            se_surv = (perc_surv * (1 - perc_surv) / n_born)^0.5,
            n_recr = sum(puprecruited),
            perc_recr = n_recr / n_born,
            se_recr = (perc_recr * (1 - perc_recr) / n_born)^0.5,
            .groups = "drop")

ggplot(surv_pred, aes(x = age)) +
  # Survival by age (fixed effect only)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, 
                  group = age_class),
              fill = "grey", alpha = 0.5) +
  geom_line(aes(y = predicted, color = age_class), 
            linewidth = 1.5) +
  # Raw survival and recruitment data
  geom_pointrange(aes(y = perc_surv,
                      ymin = perc_surv - se_surv,
                      ymax = perc_surv + se_surv),
                  repro_raw) +
  geom_pointrange(aes(y = perc_recr,
                      ymin = perc_recr - se_recr,
                      ymax = perc_recr + se_recr),
                  shape = 21,
                  filter(repro_raw, pupsex == "Female")) +
  # Add visual indicator for senescence
  geom_vline(xintercept = age_senesce - 0.5, linetype = "dashed") +
  scale_x_continuous("Age", limits = c(3.5, 18.5)) +
  scale_y_continuous("Pup survival/recruitment", 
                     limits = c(0, 0.65), 
                     labels = scales::percent) +
  scale_color_manual(values = c("firebrick", "cornflowerblue")) +
  facet_wrap(~pupsex) +
  theme_few() +
  theme(legend.position = "none",
        axis.title = element_text(size = 13))
